---
title: 'Reproducible Research: Peer Assessment 1'
author: "Joseph R. Taylor"
date: "Sunday, October 19, 2014"
output: pdf_document
subtitle: JHU/Coursera Data Science Specialization
---
  
  

### Load and Pre-Process the data

```{r}
# load the data file
thedata <- read.csv("activity.csv")
# process data to convert the date column to a date class
thedata$date <- as.Date(thedata$date)
```
  
### What is mean total number of steps taken per day?  
***Histogram the total number of steps taken each day***  
```{r}
# make a new dataframe without NA values
cleanData <- thedata[!is.na(thedata$steps), ]
# load plyr library
library(plyr)
# group values by date
groupedByDate <- ddply(cleanData, ~date, summarise, sum = sum(steps))
# construct the histogram
hist(groupedByDate$sum, xlab = "Total Steps", main = "Histogram: Total Steps 
     per Day", col = "tomato4")
```
  
***Calculate and display the mean and median total number of steps taken per 
        day***
```{r}
# Mean
mean(groupedByDate$sum)
# Median
median(groupedByDate$sum)
```
  

###What is the average daily activity pattern?###
***Make a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all days (y-axis)***

```{r}
groupedByInterval <- ddply(cleanData, ~interval, summarise, mean = mean(steps))
with(groupedByInterval, plot(interval, mean, type = "l", ylab = "Average number 
        of steps", xlab = "Interval", main = "Average daily activity", 
        col = "firebrick3"))

```

***Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?***

```{r}
# Find the maximum value
maxVal <- max(groupedByInterval$mean)
print(maxVal)
```
```{r}
# Find the line containing the maximum value
maxLine <- groupedByInterval[groupedByInterval$mean == maxVal, ]
# Find the corresponding interval
maxInterval <- maxLine$interval
print(maxInterval)
```

The maximum number of steps (on average across all the days) is 206.1698. It is 
contained in interval 835.

###Imputing Missing Values###
***Calculate and report the total number of missing values in the dataset (i.e.
the total number of rows with NAs)***

```{r}
# Calculate the sum of missing values
sum(is.na(thedata$steps))
```

***Devise a strategy for filling in all of the missing values in the dataset.***

Strategy: Five-minute interval means


***Create a new dataset that is equal to the original dataset but with the 
missing data filled in***

```{r}
# Create a new dataset
newdata <- thedata
# Locate tha NAs
missingsteps <- is.na(newdata$steps)
# Convert interval(s) to factor(s)
newdata$interval <- factor(newdata$interval)
groupedByInterval$interval <- factor(groupedByInterval$interval)

# Fill newdata on missing values wherever a missing value appears. 
# Fill missing data from groupedByInterval$mean column (steps).
newdata[missingsteps, "steps"] <- groupedByInterval[newdata[missingsteps, 
        "interval"], "mean"]
```

***Make a histogram of the total number of steps taken each day.***

***Calculate and report the mean and median total number of steps taken per 
day.***

```{r}
# group values by date
groupedByDate2 <- ddply(newdata, ~date, summarise, sum = sum(steps))
 
# construct the histogram
hist(groupedByDate2$sum, xlab = "Total Steps", main = "Histogram total steps 
     per day", col = "seagreen")
```


*Do these values differ from the estimates from the first part of the 
assignment?* 

```{r}
# Calculate mean value
mean(groupedByDate2$sum)
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;The mean does not increase.

```{r}
# Calculate median value
median(groupedByDate2$sum)
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; The median increases by one.


*What is the impact of imputing missing data on the estimates of the total 
daily number of steps?*


&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;There is negligible impact.

###Are There Differences in Activity Patterns Between Weekdays and Weekends?###

***Create a new factor variable in the dataset with two levels - "weekday" and 
"weekend" indicating whether a given date is a weekday or weekend day.***

```{r}
# Add a new column containing the day of the week
newdata$weekday = weekdays(newdata$date)
# Add a new column containing either the Weekday OR Weekend
newdata$weekday.type <- ifelse(newdata$weekday == "Saturday" | newdata$weekday
                               == "Sunday", "Weekend", "Weekday")
# Convert: column to factor
newdata$weekday.type <- factor(newdata$weekday.type)
```

***Make a panel plot containing a time series plot of the 5-minute interval 
(x-axis) and the average number of steps taken, averaged across all weekday 
days or weekend days (y-axis).***

```{r}
# Make a new dataset. Grouping data by interval and weekday.type
groupedBy.Interval.WeekDay <- ddply(newdata, ~interval + weekday.type, 
        summarise, mean = mean(steps))

# An intelligible plot 'unfactors' interval. Convert to characters
# first. Otherwise, the level values (1,2,3..) are produced.
groupedBy.Interval.WeekDay$interval <- as.numeric(as.character
        (groupedBy.Interval.WeekDay$interval))
library(lattice)
xyplot(mean ~ interval | weekday.type, groupedBy.Interval.WeekDay, type = "l", 
         layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")
```

